```{r uss-setup, include=FALSE, purl=FALSE}
#in case you forget to add a chunk label
knitr::opts_knit$set(unnamed.chunk.label = "uss-")
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache=TRUE, 
                      tidy.opts=list(width.cutoff=60), tidy=TRUE, fig.align='center', out.width='80%')
```

# Univariate models {#sec-uss-head}
\chaptermark{USS}

Start with one paragraph overview.

A script with all the R code in the chapter can be downloaded  [here](./Rcode/template.R). The Rmd for this chapter can be downloaded [here](./Rmds/template.Rmd).

### Data and packages {-}

All the data used in the chapter are in the **MARSS** package.  Install the package, if needed, and load to run the code in the chapter.
```{r uss-load-packages, results='hide', message=FALSE, warnings=FALSE}
library(MARSS)
```

## Introduction {#sec-uss-intro}

Introdution. Here's how to do an equation with numbering.

\begin{equation}
\begin{gathered}
\mathbf{y}_t = \mathbf{Z}\mathbf{x}_t+\mathbf{a}+\mathbf{v}_t \text{ where } \mathbf{v}_t \sim \text{MVN}(0,\mathbf{R}) \\
\mathbf{x}_t = \mathbf{x}_{t-1}+\mathbf{w}_t \text{ where } \mathbf{w}_t \sim \text{MVN}(0,\mathbf{Q}) \\
\end{gathered}
(\#eq:template1)
\end{equation}

```{r child=file.path(here::here(), "chapters", "chap02-uss", "sec2.Rmd")}
```

## Section 3 {#sec-uss-3}

Another section.

## Discussion  {#sec-uss-discussion}

Here is a discussion.
