## New section {#sec-uss-2}

New section. Example of an equation with matrix.

\begin{equation}
\begin{bmatrix}
    y_{1} \\
    y_{2} \\
    y_{3} \\
    y_{4} \\
    y_{5} \end{bmatrix}_t = 
 \begin{bmatrix}
    z_{11}&z_{12}&z_{13}\\
    z_{21}&z_{22}&z_{23}\\
    z_{31}&z_{32}&z_{33}\\
    z_{41}&z_{42}&z_{43}\\
    z_{51}&z_{52}&z_{53}\end{bmatrix}   
 \begin{bmatrix}
    x_{1} \\
    x_{2} \\
    x_{3} \end{bmatrix}_t + 
 \begin{bmatrix}
    a_1 \\
    a_2 \\
    a_3 \\
    a_4 \\
    a_5 \end{bmatrix} + 
 \begin{bmatrix}
    v_{1} \\
    v_{2} \\
    v_{3} \\
    v_{4} \\
    v_{5} \end{bmatrix}_t.
(\#eq:template2-y)
\end{equation}

and the process model would look like

\begin{equation}
\begin{bmatrix}
    x_{1} \\
    x_{2} \\
    x_{3} \end{bmatrix}_t = 
 \begin{bmatrix}
    1&0&0 \\
    0&1&0 \\
    0&0&1 \end{bmatrix} 
 \begin{bmatrix}
    x_{1}\\
    x_{2}\\
    x_{3}\end{bmatrix}_{t-1} +
 \begin{bmatrix}
    w_{1} \\
    w_{2} \\
    w_{3} \end{bmatrix}_t 
(\#eq:tempate2-x)
\end{equation}


The observation errors would be

\begin{equation}
\begin{bmatrix}
    v_{1} \\
    v_{2} \\
    v_{3} \\
    v_{4} \\
    v_{5} \end{bmatrix}_t 
 \sim \text{MVN} \begin{pmatrix}
    \begin{bmatrix}
    0 \\
    0 \\
    0 \\
    0 \\
    0 \end{bmatrix},
 \begin{bmatrix}
    r_{11}&r_{12}&r_{13}&r_{14}&r_{15}\\
    r_{12}&r_{22}&r_{23}&r_{24}&r_{25}\\
    r_{13}&r_{23}&r_{33}&r_{34}&r_{35}\\
    r_{14}&r_{24}&r_{34}&r_{44}&r_{45}\\
    r_{15}&r_{25}&r_{35}&r_{45}&r_{55}\end{bmatrix}
\end{pmatrix}
(\#eq:template2-oe)
\end{equation}

And the process errors would be

\begin{equation}
\begin{bmatrix}
    w_{1} \\
    w_{2} \\
    w_{3} \end{bmatrix}_t
\sim \text{MVN} \begin{pmatrix}
 \begin{bmatrix}
    0 \\
    0 \\
    0 \end{bmatrix},
 \begin{bmatrix}
    q_{11}&q_{12}&q_{13}\\
    q_{12}&q_{22}&q_{23}\\
    q_{13}&q_{23}&q_{33}\end{bmatrix}
\end{pmatrix}.
(\#eq:template2-pe)
\end{equation}


And here is a plot.

```{r uss-plot100}
plot(1:100)
```

```{r uss-plot10}
plot(1:10)
```