---
title: "Chapter 3"
output:
  github_document:
    html_preview: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This chapter shows how you might use Target Markdown to create a target pipeline for a specific chapter. It will create the {target} files in the chapter folder so you'll want a separate folder for the chapter. To work on the chapter, you would set the working directory to the chapter folder. Then you can use the usual {targets} workflow and visualizations. 

The chapter at https://books.ropensci.org/targets/markdown.html walks through it in detail. This R Markdown template is based off what you get using File > New > RMarkdown > Template > Targets Markdown

## How to run this.

Just hit the Knit button.

What's it going to do? Basically it will create the target objects in `_targets` and then the markdown text will use those target objects. That way you don't end up re-running lengthy things. How is this different from caching? Well you explicitly specify the dependencies of your objects. {targets} is smart and will figure out if it needs to re-make an object based on whether it or its dependencies changed. Also caching crashes {bookdown} constantly.

## Packages

The example requires several R packages, and `targets` must be version 0.5.0.9000 or above. 

```{r, eval = FALSE}
install.packages(c("biglm", "dplyr", "ggplot2", "readr", "targets", "tidyr"))
```

## Setup

If you are using old versions of `targets` (<= 0.7.0) and/or `knitr` (<= 1.33), you will need to load the `targets` package in the R Markdown document in order for Target Markdown code chunks to work.

```{r}
library(targets)
```

Near the top of the document, you may also wish to remove the `_targets_r` directory previously written by non-interactive runs of the report. Otherwise, your pipeline may contain superfluous targets.

```{r}
library(targets)
tar_unscript()
```

## Globals

We first define some global options/functions common to all targets. The function below plots a histogram of ozone concentrations, and our histogram target will need it.

```{targets example-globals, tar_globals = TRUE}
options(tidyverse.quiet = TRUE)
tar_option_set(packages = c("biglm", "dplyr", "ggplot2", "readr", "tidyr"))
create_plot <- function(data) {
  ggplot(data) +
    geom_histogram(aes(x = Ozone), bins = 12) +
    theme_gray(24)
}
```

## Targets

This is the important part. Notice that it uses `{targets ...}` not `{r ...}`. You need to define dependencies (like data) as a target object so that {targets} can keep track of whether it needs to update objects based on changes to its dependencies.

*How will you use these?* Not here. This is just defining them. Scroll down to see how to use them. Note you don't have to define all your targets in one place. Scroll down to see examples.

Our first target borrows the `airquality` dataset built into base R.

```{targets raw-data}
tar_target(raw_data, airquality)
```

Our next targets preprocess the data, make a histogram, and fit a model.

```{targets downstream-targets}
list(
  tar_target(data, raw_data %>% filter(!is.na(Ozone))),
  tar_target(hist, create_plot(data))
)
```

Set the `tar_simple` chunk option to `TRUE` to define a single target with the command in the code chunk. The chunk below only contains `biglm(Ozone ~ Wind + Temp, data)` in the source, but because `tar_simple` is `TRUE`, it is shorthand for `tar_target(name = fit, command = biglm(Ozone ~ Wind + Temp, data))`. All other arguments to `tar_target()` are set to their default values (configurable with `tar_option_set()`).

```{targets fit, tar_simple = TRUE}
biglm(Ozone ~ -1 + Wind + Temp, data)
```

You need to be careful how you name things. You don't want to name an object to the name of a targets object. So like in the code above, you don't want to use `data` as a name in that code chunk since `data` is now a targets object.

## Pipeline

If you ran all the `{targets}` chunks in non-interactive mode, then your R scripts are set up to run the pipeline. *I don't know what this means.*

```{r}
tar_make()
```

## Output

You can retrieve results from the `_targets/` data store using `tar_read()` or `tar_load()`.

```{r, message = FALSE}
library(biglm)
tar_read(fit)
```

```{r}
tar_read(hist)
```



